Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus in a Turkey B-cell Line

### 2024-05-06 manuscript_thev_transcriptome.tex
first round changes after jvi review

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses,
causing infection in virtually all types of vertebrates studied to date.
Their double-stranded linear DNA genomes range between 26 and 45kb in
size, producing a broad repertoire of transcripts via highly complex
alternative splicing patterns (1, 2). The AdV genome is one of the most
optimally economized; both the forward and reverse DNA strands harbor
protein-coding genes, making it highly gene-dense. There are 16 genes
termed ``genus-common'' that are homologous in all AdVs; these are
thought to be inherited from a common ancestor. All other genes are
termed ``genus-specific''. The genus-specific genes tend to be located
at the termini of the genome while genus-common genes are usually
towards the center of the genome (1). This pattern is also observed in
\emph{Poxviridae} and \emph{Herpesviridae}, which also have linear DNA
genomes (1, 3, 4). The family \emph{Adenoviridae} consists of five
genera: \emph{Mastadenovirus} (MAdV), \emph{Aviadenovirus},
\emph{Atadenovirus}, \emph{Ichtadenovirus}, and \emph{Siadenovirus}
(SiAdV) to which turkey adenovirus 3 also called turkey hemorrhagic
enteritis virus (THEV) belongs (5--10). Members of SiAdV have the
smallest genome size (\textasciitilde26 kb) and gene content of all
known AdVs (see \textbf{Figure 1}) (1, 2, 6).

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV
both infect turkeys, with THEV-V causing hemorrhagic enteritis (HE), a
debilitating acute disease predominantly affecting turkey poults
characterized by immunosuppression (IS), intestinal lesions leading to
bloody diarrhea, and up to 80\% mortality (2, 11--13). While the current
vaccine strain (a THEV-A strain isolated from a pheasant; Virginia
Avirulent Strain {[}VAS{]}) has proven effective at preventing HE in
turkey poults, it still retains its immunosuppressive ability. Thus,
vaccinated birds are rendered more susceptible to opportunistic
infections and death than unvaccinated birds leading to substantial
economic losses (11, 14--16). To eliminate the immunossupressive effect
of the vaccine strain, a thorough investigation of the culprit viral
genes mediating this phenomenon is essential. However, the transcriptome
(splicing and gene expression patterns) of THEV has not been
characterized, making an investigation of specific IS-related viral
genes impractical.

A myriad of studies have elucidated the AdV transcriptome in fine detail
(17, 18). However, a large preponderance of studies focus on MAdVs --
specifically human AdVs. Thus, most of the current knowledge regarding
AdV gene expression and replication is based on MAdV studies, which is
generalized for all other AdVs (10, 19). MAdV transcription is
temporally regulated; therefore, genes are categorized into five early
transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM)
units (pIX and IVa2), and one major late transcription unit (MLTU or
major late promoter {[}MLP{]} region), which generates five families of
late mRNAs (L1-L5) based on the polyadenylation site. An additional gene
(UXP or U exon) is located on the reverse strand. The early genes encode
non-structural proteins such as enzymes or host-cell modulating
proteins, primarily involved in DNA replication, or providing the
necessary intracellular niche for optimal replication while late genes
encode structural proteins that act as capsid proteins, promote virion
assembly, or direct genome packaging. The immediate early genes E1A are
expressed first, followed by the delayed early genes, E1B, E2, E3 and
E4. Then the intermediate early genes, IVa2 and pIX are expressed
followed by the late genes (10, 17, 18). It is noteworthy that the MLP
shows basal transcriptional activity during early infection (before DNA
replication), with a comparable efficiency to other early viral
promoters, but it reaches its maximal activity during late infection
(after DNA replication). However, during early infection only a subset
of the MLP-derived transcripts are expressed (10). MAdV makes an
extensive use of alternative RNA splicing to produce a very complex
array of mRNAs. All but the pIX mRNA undergo at least one splicing
event. For instance, the MLTU produces over 20 distinct splice variants
all of which contain three non-coding exons at the 5'-end (collectively
known as the tripartite leader; TPL) (17, 18). There is also an
alternate three-exon 5' non-coding leader sequence present in varying
amounts on a subset of MLTU mRNAs (known as the x-, y-, and z-leaders).
Lastly, there is the i-leader exon, which is infrequently included
between the second and third TPL exons, and codes for the i-leader
protein (20). Thus, the MLTU produces a complex repertoire of mRNA with
diverse 5' untranslated regions (UTRs) spliced onto different 3' coding
exons which are grouped into five different 3'-end classes (L1-L5) based
on polyadenylation site. Each transcription unit (TU) contains its own
promoter driving the expression of the array of mRNA transcripts
produced via alternative splicing in the unit (10, 17, 18). The
promoters are activated at different phases of the infection by proteins
from previously activated TUs. Paradoxically, the early-to-late phase
transition during infection requires the L4 gene products, 22K and 33K,
which should only be available after the transition. However, a promoter
in the L4 region (L4P) that directs the expression of these two proteins
independent of the MLP was found, resolving the paradox (10, 17, 21).
During translation of AdV mRNA, recent studies using long-read direct
RNA sequencing strongly suggest the potential usage of secondary start
codons; adding to what was already a highly complex system for gene
expression (17, 22).

@@

\textbf{Overview of sequencing data and analysis pipeline outputs}\\
A previous study by Aboezz \emph{et al} showed that almost all THEV
transcripts were detectable beginning at 4 hours, with a full
replication cycle of about 18-hours post infection (hpi) (24).
Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and
72-hpi to ensure a wide time window to sample all transcripts. Our
paired-end RNA sequencing (RNA-seq) experiment yielded an average of
107.1 million total reads of 149 bp in length per time-point, which were
simultaneously mapped to both the virus (THEV) and host (\emph{Meleagris
gallopavo}) genomes using the Hisat2 (25) reference-based aligner. A
total of 18.1 million reads from all time-points mapped to the virus
genome; this provided good coverage/depth, leaving no regions unmapped.
The mapped reads to the virus genome increased substantially from a mere
432 reads at 4 hpi to 16.9 million reads at 72 hpi (\textbf{Table 1},
\textbf{Figure 2A}). From the mapped reads, we identified a total of
2,457 unique THEV splice junctions from all time-points, with splice
junctions from the later time-points being supported by significantly
more sequence reads than earlier time-points. For example, all the 13
unique junctions at 4 hpi had less than 10 reads supporting each one,
averaging a mere 2.8 reads per junction. Conversely, the 2374 unique
junctions at 72 hpi averaged 898.4 reads per junction, some junctions
having coverage as high as 322,677 reads. The substantial increases in
splice junction and mapping reads to the THEV genome over time denotes
an active infection and a successful viral replication, and correlates
with our quantitative PCR (qPCR) assay quantifying the total number of
viral genome copies over time (\textbf{Figure 2B}).

Using StringTie (25), we assembled the data into potential transcripts
using the genomic location of the prior predicted THEV ORFs as a guide.
In the consolidated transcriptome, a composite of all non-redundant
transcripts from all time points, we counted a total of 29 novel
transcripts. We found that a subset of exons in the viral transcripts
match the predicted ORFs exactly, with the majority of the exons being
longer and spanning multiple predicted ORFs (\textbf{Figure 3}).

@@

\textbf{Changes in THEV splicing profile over time}\\
AdV gene expression occurs under exquisite temporal control with each
promoter typically producing one or few pre-mRNAs that undergo
alternative splicing to yield a repertoire of mature mRNAs. To evaluate
the activity of each promoter over time, StringTie and Ballgown (a
program for statistical analysis of assembled transcriptomes) (26) were
used to estimate the normalized expression levels of all transcripts for
each time point in Fragments Per Kilobase of transcript per Million
mapped reads (FPKM) units. Very few unique splice junctions, reads, and
transcripts were counted at 4 hpi; hence, this time point was excluded
in this analysis.

Considering individual mRNAs, TRXPT\_21 -- from the E2 region -- was the
most significantly expressed at 12 hpi, constituting 33.58\% of the
total expression of all transcripts. Transcripts in the E3 and E4
regions also contributed significant proportions, and noticeably, some
MLP region transcripts. The later time points were dominated by the MLP
region transcripts -- TRXPT\_10 and TRXPT\_14 were the most abundantly
expressed at 24 and 72 hpi, respectively, as expected (\textbf{Figure
4A}). When we performed analysis of the FPKM values of transcripts per
region we found a similar pattern: the E2 region was the most abundantly
expressed at 12 hpi, after which the MLP region assumes predominance
(\textbf{Figure 4B}). Secondly, we estimated relative abundances of all
splice junctions at each time point using the raw reads. Only junctions
with a read coverage of at least 1\% of the total splice junction reads
at the given time point were considered significant and included in
\textbf{Tables 2a-2c}. At 12 hpi, 18 junctions meet the 1\% threshold,
and were comprised of predominantly early region (E1, E2, E3, and E4)
junctions, albeit the MLTU was the single most preponderant region
overall, constituting 38.8\% of all the junction reads (\textbf{Table
2a} and \textbf{Supplementary Table S1a}). The most abundant junctions
at 12 hpi remained the most significantly expressed at 24 hpi also.
However, here, the MLP-derived junctions were unsurprisingly even more
preponderant overall, accounting for 45.7\% of all the junction reads
counted (\textbf{Table 2b} and \textbf{Supplementary Table S1b}). At 72
hpi, the trend of increased activity of the MLP continued as expected;
at this time, the MLP region junctions were not only the most abundant
overall -- accounting for 67.3\% of all junction reads, -- but also
contained the most significantly expressed individual junctions
(\textbf{Table 2c}, \textbf{Supplementary Table S1c} and \textbf{Figure
4C}). When we limited this analysis to only junctions in the final
transcriptome, we observed the relative abundances of the junctions for
each region over time to be similar to the pattern seen with all the
junctions included (\textbf{Figure 4D}).

We also analyzed splice donor and acceptor site nucleotide usage over
time to investigate any peculiarities that THEV may show, generally or
over the course of the infection. We found that most splice
donor-acceptor sequences were unsurprisingly the canonical GU-AG
nucleotides. However, the splice acceptor-donor pairing became less
specific over time, such that all combinations of nucleotide pairs were
eventually detected (\textbf{Figure 5}).

\textbf{Early Region 1 (E1) transcripts}\\
This region in MAdVs is the first transcribed after successful entry of
the viral DNA into the host cell nucleus, albeit at low levels (18). The
host transcription machinery solely mediates the transcription of this
region. After their translation, the E1 proteins in concert with a
myriad of host transcription factors activate other viral promoters
(10). In MAdVs, this region is subdivided into E1a and E2b units, but
the transcripts found in our data categorized under this region do not
appear to be subdivided.

Only two ORFs (ORF1 {[}sialidase{]} and Hyd) are predicted in this
region; however, we discovered four novel transcripts in this region,
which collectively contain 3 unique splice junctions (\textbf{Figure
6}). Most of the ORFs of the novel transcripts are distinct from the
predicted ORFs, but they all have the coding potential (CP) for the
predicted Hyd protein as the 3'-most coding sequence (CDS) if secondary
start codon usage is considered as reported for other AdVs (17, 18). The
5'-most CDS of TRXPT\_1 is multi-exonic, encoding a novel 17.9
kilodalton (kDa), 160 residue {[}amino acids (aa){]} protein (ORF9).
From its 5'-most start codon (SC), TRXPT\_2 encodes the largest protein
in this region -- a 64.3 kDa, 580 aa protein (ORF10) with the same SC as
ORF9 (position 211 bp). ORF10 spans almost the entire predicted ORF1 and
Hyd, coming short in two regards: it is spliced from 1655 bp to 1964 bp
(ORF1's C-terminus, including the stop codon), and its stop codon (STC;
position 2312) is 13 bp short of the Hyd STC. However, it has an SC 102
bp upstream and in-frame with ORF1's predicted SC. Thus, ORF10 shares
substantial protein sequence similarity with ORF1 but not with Hyd, as
the SC of Hyd is not in-frame. Without its splice site removing the ORF1
STC, TRXPT\_2 would encode a longer variant of ORF1, starting from an
upstream SC. TRXPT\_3 is almost identical to TRXPT\_1, except for the
lack of TRXPT\_1's second exon. Our RNA-seq data show that all E1
transcripts share the same transcription termination site (TTS; at
position 2325 bp). However, TRXPT\_3 and TRXPT\_4 seem to have
transcription start sites (TSS) downstream of the TSS of TRXPT\_1 and
TRXPT\_2 (E1 TSS; position: 54 bp). Given that studies in MAdVs show
that E1 mRNAs share not only a common TTS but also the TSS, and only
differ from each other regarding the internal splicing (18), it is
likely that TRXPT\_3 and TRXPT\_4 are incomplete, and their actual TSS
just like the TTS are identical for all E1 transcripts. Regardless of
the TSS considered for TRXPT\_3, the CP remains unaffected. Its 5'-most
CDS, beginning at 1965 bp and sharing the same STC as ORF9, produces a
13.1 kDa, 115 residue protein (ORF4). ORF4 was predicted in an earlier
study (27) but was excluded in later studies (1, 12); however, our data
suggest it is a genuinely expressed ORF. Unlike TRXPT\_3, the CP of
TRXPT\_4 is affected by the TSS considered; if we consider its
unmodified TSS, then its CP is identical to TRXPT\_3. However, if we
assume that TRXPT\_4 uses the E1 TSS, then the 5'-most CDS is a
distinct, novel, multi-exonic 15.9 kDa, 143 aa protein (ORF11) with the
same SC as ORF9 and ORF10 but with a unique STC.

The splice junctions of all transcripts in this region (except the
junction for TRXPT\_4) were validated by cloning of viral cDNA and
Sanger sequencing (\textbf{Supplementary PCR methods}). During the
validation of TRXPT\_2, we found ORF1 to be present on the agarose gel
(an unspliced band size) and Sanger sequencing results showed it to be a
transcribed mRNA (\textbf{Supplementary PCR methods}). This was
corroborated by our 3' RACE experiment, which showed a transcript
(TRXPT\_2B) spanning the entire ORF1 and Hyd ORFs without any splicing,
with a poly-A tail immediately after the E1 TTS. The 5'-most CDS of this
transcript (TRXPT\_2B) would encode ORF1. However, TRXPT\_2B has an
upstream and in-frame SC to the predicted SC of ORF1, suggesting that
the predicted ORF1 CDS is truncated -- the actual ORF1 (eORF1) that is
expressed shares the same SC as ORF10, but has a unique STC.

\textbf{Early Region 2 (E2) and Intermediate Region (IM) transcripts}\\
The E2 TU expressed on the anti-sense strand is subdivided into E2A and
E2B and encodes three classical AdV proteins -- pTP and Ad-pol (E2B
proteins), and DBP (E2A protein) -- essential for genome replication
(17, 18). Unlike MAdV where two promoters (E2-early and E2-late) are
known (17), we discovered only a single TSS (E2 TSS; 18,751 bp) from
which both E2A and E2B transcription is initiated. However, similar to
MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B
transcripts share the TTS of the IVa2 transcript of the IM region (17,
18) (\textbf{Figure 7}).

The E2A ORF, DBP is one of three THEV ORFs predicted to be spliced from
two exons. The corresponding transcript (TRXPT\_21) found in our data
matches this predicted splice junction precisely but with a non-coding
additional exon at the 5'-end (E2-5'UTR) at position 18,684-18,751 bp.
Thus, TRXPT\_21 is a three-exon transcript encoding DBP (380 residues,
43.3 kDa) precisely as predicted. This transcript (TRXPT\_21) was also
corroborated in a 3' RACE experiment. Additionally, from the 3' RACE
data, we found a splice variant of TRXPT\_21 which retains the second
intron leading to a 2-exon transcript. This new transcript (TRXPT\_21B),
albeit longer due to retaining the second intron and possessing a short
3' UTR, encodes a truncated isoform of DBP (tDBP) because the SC
utilized by TRXPT\_21, is followed shortly by STCs in the retained
intron. The SC 173 bp downstream of the DBP SC yields tDBP (a 346
residue, 39.3 kDa product), which is in-frame of DBP but entirely
contained in the second exon. TRXPT\_21 and TRXPT\_21B share a common
TTS but TRXPT\_21B as seen in our 3'-RACE data, extends 39 bp into an
adenine/thymine (A/T)-rich sequence before the poly-A tail sequence
occurs, suggesting this position (16,934 bp) as the true E2A TTS
(\textbf{Figure 7}).

The E2B region transcripts also start with the E2-5'UTR but extend
thousands of base pairs downstream to reach the TTS at 2334 bp in the IM
region, which is immediately followed by an A/T-rich sequence (position
2323-2339 bp) where polyadenylation probably occurs. Interestingly, the
TTS of the E1 region (position 2,325 bp) on the sense strand is also in
the immediate vicinity of this A/T-rich sequence, which is almost
palindromic; hence it likely serves as the polyadenylation signal for
both E1 and E2B/IM transcripts. The E2B transcripts TRXPT\_6 and
TRXPT\_7 are almost identical except for an extra splice junction at the
3'-end of TRXPT\_6, making TRXPT\_6 a five-exon transcript and TRXPT\_7,
four exons (\textbf{Figure 7}). TRXPT\_7 has the CP for both classical
proteins (pTP and Ad-pol) encoded in this region, of which the pTP ORF
is predicted to be spliced from two exons just like in all other AdVs.
The predicted splice junction of pTP is corroborated by our data;
however, the full transcript is markedly longer than the predicted ORF:
there are two novel non-coding 5' exons, the third exon (containing the
SC of pTP) is significantly longer than predicted, and the last exon
containing the bulk of the CDS is more than triple the predicted size of
pTP. The first two exons are 5'-UTRs because the SC here is immediately
followed by STCs; thus, the 5'-most SC (position 10,995 bp) of the third
exon which matches the predicted SC of pTP is utilized. The encoded
product is identical to the predicted pTP protein (597 residues; 70.5
kDa). If secondary SC (secSC) usage is considered, with SC at 6768 bp
and STC at 3430 bp, the encoded product is identical to the predicted
Ad-pol (polymerase) protein (1112 residues; 129.2 kDa). TRXPT\_6 differs
from TRXPT\_7 by containing an extra splice site at 3447-3515 bp.
However, the CP remains similar to that of TRXPT\_7 except the Ad-pol
encoded from the secSC is a truncated isoform with a new STC resulting
from the splice site.

While both TRXPT\_6 and TRXPT\_7 have the CP for Ad-pol with secSC
usage, in all AdVs studied, the two proteins (pTP and Ad-pol) are
encoded by separate mRNAs with identical first three 5' exons and TTS,
but the splice junction to the terminal exons are different. We checked
for a longer splice junction between the third and fourth (terminal)
exons of TRXPT\_7 with our junction validation method (targeted PCR,
cloning, and Sanger sequencing) and discovered a unique splice junction
(10,981-7062 bp) not found in our RNA-seq data. If initiated from the E2
TSS and terminated at the E2 TTS, this transcript (TRXPT\_31) would
encode Ad-pol exactly as predicted as its 5'-most CDS (\textbf{Figure
7}).

@@

>>>>

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses,
causing infection in virtually all types of vertebrates studied to date.
Their double-stranded linear DNA genomes range between 26 and 45kb in
size, producing a broad repertoire of transcripts via highly complex
alternative splicing patterns (1, 2). The AdV genome is one of the most
optimally economized; both the forward and reverse DNA strands harbor
protein-coding genes, making it highly gene-dense. There are 16 genes
termed ``genus-common'' that are homologous in all AdVs, presumably
inherited from a common ancestor. All other genes are termed
``genus-specific''. The genus-specific genes tend to be located at the
termini of the genome while genus-common genes are usually towards the
center of the genome (1). This pattern is also observed in
\emph{Poxviridae} and \emph{Herpesviridae}, which also have linear DNA
genomes (1, 3, 4). The family \emph{Adenoviridae} consists of five
genera: \emph{Mastadenovirus} (MAdV), \emph{Aviadenovirus},
\emph{Atadenovirus}, \emph{Ichtadenovirus}, and \emph{Siadenovirus}
(SiAdV) to which turkey adenovirus 3 also called turkey hemorrhagic
enteritis virus (THEV) belongs (5--10). Members of SiAdV have the
smallest genome size (\textasciitilde26 kb) and gene content of all
known AdVs (see \textbf{Figure 1}) (1, 2, 6).

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV
both infect turkeys, with THEV-V causing hemorrhagic enteritis (HE), a
debilitating acute disease predominantly affecting turkey poults
characterized by immunosuppression, intestinal lesions leading to bloody
diarrhea, and up to 80\% mortality (2, 11--13). While the current
vaccine strain (a THEV-A called Virginia Avirulent Strain {[}VAS{]}) has
proven effective at preventing HE in turkey poults, it still retains its
immunosuppressive ability. Thus, vaccinated birds are rendered more
susceptible to opportunistic infections and death than unvaccinated
birds leading to substantial economic losses (11, 14--16). To eliminate
the immunosupressive effect of the vaccine strain, a thorough
investigation of the culprit viral genes mediating this phenomenon is
essential. However, the transcriptome (splicing and gene expression
patterns) of THEV has not been characterized, making an investigation of
specific immunosuppressive viral genes impractical.

A myriad of studies have elucidated the AdV transcriptome in fine detail
(17, 18). However, a large preponderance of studies focus on MAdVs --
specifically human AdVs. Thus, most of the current AdV gene expression
and replication knowledge is based on MAdV studies, which is generalized
for all other AdVs (10, 19). MAdV transcription is temporally regulated;
their genes are categorized into five early transcription units (E1A,
E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and
one major late transcription unit (MLTU or major late promoter {[}MLP{]}
region), which generates five families of late mRNAs (L1-L5) based on
the polyadenylation site. An additional gene (UXP or U exon) is located
on the reverse strand. The early genes encode non-structural proteins
such as enzymes or host-cell modulating proteins, primarily involved in
DNA replication, or providing the necessary intracellular niche for
optimal replication while late genes encode structural proteins that act
as capsid proteins, promote virion assembly, or direct genome packaging.
The immediate early genes E1A are expressed first, followed by the
delayed early genes, E1B, E2, E3 and E4. Then the intermediate early
genes, IVa2 and pIX are expressed followed by the late genes (10, 17,
18). It is noteworthy that the MLP shows basal transcriptional activity
during early infection (before DNA replication), with a comparable
efficiency to other early viral promoters, but it reaches its maximal
activity during late infection (after DNA replication). However, during
early infection only a subset of the MLP-derived transcripts are
expressed (10). MAdV makes an extensive use of alternative RNA splicing
to produce a very complex array of mRNAs. All but the pIX mRNA undergo
at least one splicing event. For instance, the MLTU produces over 20
distinct splice variants all containing three non-coding exons at the
5'-end (collectively known as the tripartite leader; TPL) (17, 18).
There is also an alternate three-exon 5' non-coding leader sequence
present in varying amounts on a subset of MLTU mRNAs (known as the x-,
y-, and z-leaders). Lastly, there is the i-leader exon, which is
infrequently included between the second and third TPL exons, and codes
for the i-leader protein (20). Thus, the MLTU produces a complex
repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced
onto different 3' coding exons which are grouped into five different
3'-end classes (L1-L5) based on polyadenylation site. Each transcription
unit (TU) contains its own promoter driving the expression of the array
of mRNA transcripts produced via alternative splicing in the unit (10,
17, 18). The promoters are activated at different phases of the
infection by proteins from previously activated TUs. Paradoxically, the
early-to-late phase transition during infection requires the L4 gene
products, 22K and 33K, which should only be available after the
transition. However, a promoter in the L4 region (L4P) that directs the
expression of these two proteins independent of the MLP was found,
resolving the paradox (10, 17, 21). During translation of AdV mRNA,
recent studies using long-read direct RNA sequencing strongly suggest
the potential usage of secondary start codons; adding to what was
already a highly complex system for gene expression (17, 22).

@@

\textbf{Overview of sequencing data and analysis pipeline outputs}\\
A prior study by Aboezz \emph{et al}. demonstrated that nearly all THEV
transcripts became detectable starting at 4 hours post-infection (hpi),
with one replication cycle concluding around 18 hpi (24). Consequently,
we harvested infected MDTC-RP19 cells (MOI of 100 genome copy
numbers/cell) at 4-, 12-, 24-, and 72-hpi to capture all transcripts
within a broad time window. Our paired-end RNA sequencing (RNA-seq)
experiment generated an average of 107.1 million total reads of 149 bp
length per time-point. These reads were concurrently mapped to both the
virus (THEV) and host (\emph{Meleagris gallopavo}) genomes using the
Hisat2 (25) reference-based aligner. A total of 18.1 million reads from
all time-points mapped to the virus genome, providing comprehensive
coverage and leaving no regions unmapped. The mapped reads to the virus
genome increased significantly from a scant 432 reads at 4 hpi to 16.9
million reads at 72 hpi (\textbf{Table 1}, \textbf{Figure 2A}). From
these mapped reads, we identified 2,457 unique THEV splice junctions
across all time-points, with later time-points exhibiting significantly
more sequence reads supporting the splice junctions than earlier
time-points. For instance, all 13 unique junctions at 4 hpi had fewer
than 10 supporting reads each, averaging only 2.8 reads per junction. In
contrast, the 2374 unique junctions at 72 hpi averaged 898.4 reads per
junction, with some junctions reaching as high as 322,677 reads. The
marked increase in splice junction and mapping reads to the THEV genome
over time indicates an active infection and successful viral
replication, which is corroborated by our quantitative PCR (qPCR) assay
that quantified the total number of viral genome copies over time
(\textbf{Figure 2B}).

Using StringTie (25), we assembled the data into potential transcripts,
guided by the genomic locations of the previously predicted THEV ORFs.
In the consolidated transcriptome, a composite of all non-redundant
transcripts across all time points, we identified a total of 29 novel
transcripts. We found that a subset of exons in the viral transcripts
match some predicted ORFs exactly, with the majority of the exons being
longer and spanning multiple predicted ORFs (\textbf{Figure 3}).

@@

\textbf{Changes in THEV splicing profile over time}\\
AdV gene expression is subject to meticulous temporal regulation, with
each promoter typically generating one or a few pre-mRNAs. These
pre-mRNAs undergo alternative splicing to produce a diverse array of
mature mRNAs. To assess the temporal activity of each promoter, we
utilized StringTie and Ballgown (a tool for statistical analysis of
assembled transcriptomes) (26). These tools estimated the normalized
expression levels of all transcripts at each time point, measured in
Fragments Per Kilobase of transcript per Million mapped reads (FPKM)
units. At 4 hpi, we counted very few unique splice junctions, reads, and
transcripts; hence, this time-point was excluded from this analysis.

Examining individual mRNAs, TRXPT\_21 -- from the E2 region -- was the
most significantly expressed at 12 hpi, constituting 33.58\% of the
total expression of all transcripts. Transcripts in the E3 and E4
regions also contributed substantial proportions, along with some MLP
region transcripts. The later time points were dominated by the MLP
region transcripts --- TRXPT\_10 and TRXPT\_14 were the most abundantly
expressed at 24 and 72 hpi, respectively (\textbf{Figure 4A}). Our
analysis of the FPKM values of transcripts per region/TU revealed a
similar pattern: the E2 region was the most abundantly expressed at 12
hpi, after which the MLP region assumed dominance (\textbf{Figure 4B}).

Next, we estimated the relative abundances of all splice junctions at
each time point using the raw reads. Only junctions with a read coverage
of at least 1\% of the total splice junction reads at the given time
point were considered significant and included in \textbf{Tables 2a-2c}.
At 12 hpi, 18 junctions met the 1\% threshold, predominantly from early
regions (E1, E2, E3, and E4), although the MLTU was the single most
predominant region overall, constituting 38.8\% of all the junction
reads (\textbf{Table 2a} and \textbf{Supplementary Table S1a}). The most
abundant junctions at 12 hpi remained the most significantly expressed
at 24 hpi. However, here, the MLP-derived junctions unsurprisingly
became even more predominant overall, accounting for 45.7\% of all the
junction reads counted (\textbf{Table 2b} and \textbf{Supplementary
Table S1b}). At 72 hpi, the trend of increased activity of the MLP
continued as expected; at this time, the MLP region junctions were not
only the most abundant overall --- accounting for 67.3\% of all junction
reads, -\/-- but also contained the most significantly expressed
individual junctions (\textbf{Table 2c}, \textbf{Supplementary Table
S1c} and \textbf{Figure 4C}). When we limited this analysis to only
junctions in the final transcriptome, we observed the relative
abundances of the junctions for each region over time to be similar to
the pattern seen with all the junctions included (\textbf{Figure 4D}).

Finally, we analyzed splice donor and acceptor site nucleotide usage
over time to investigate any peculiarities that THEV may exhibit,
generally or over the course of the infection. We found that most splice
donor-acceptor sequences were, unsurprisingly, the canonical GU-AG
nucleotides. However, the splice acceptor-donor pairing became less
specific over time, such that all combinations of nucleotide pairs were
eventually detected (\textbf{Figure 5}).

\textbf{Early Region 1 (E1) transcripts}\\
In MAdVs, E1 is the first region transcribed post-viral DNA entry into
the host cell nucleus, mediated solely by host transcription machinery
(18). Translated E1 proteins subsequently activate other viral promoters
in conjunction with host transcription factors (10). Despite subdivision
into E1a and E1b units in MAdVs, our THEV data does not reflect this.
This region is predicted to encode only two ORFs: ORF1 (sialidase) and
Hyd (a hydrophobic product with unknown function) in THEV.

We identified four novel transcripts in this region, containing 3 unique
splice junctions (\textbf{Figure 6}), and encoding four distinct novel
ORFs in addition to Hyd. All transcripts have coding potential (CP) for
the Hyd protein as the 3'-most coding sequence if secondary start codon
usage is considered (17, 18). Also, all the transcripts have a common
transcription termination site (TTS; at position 2325 bp), but TRXPT\_1
and TRXPT\_2 have an upstream transcription start sites (TSS) to
TRXPT\_3 and TRXPT\_4. Given that E1 mRNAs in MAdVs share a common TTS
and TSS, differing only in the internal splicing (18), we consider the
upstream TSS (position 54 bp) as the actual TSS for all E1 region
transcripts. We also identified the canonical polyadenylation signal
(PAS; AAUAAA) in the immediate context of the TTS at position 2323 bp
(location of the ``U'' in the PAS sequence); see \textbf{Supplementary
Table S2}.

From the 5'-most start codon (SC), TRXPT\_1 encodes a multi-exonic novel
17.9 kDa, 160 residue protein (ORF9). TRXPT\_2 encodes a two-exon 66.4
kDa, 597 residue novel protein (ORF10), spanning almost the entire
predicted ORF1 and Hyd. The intron of TRXPT\_2 excludes the C-terminus
of ORF1 (including its stop codon) from ORF10 but the SC of ORF10 is 102
bp upstream and in-frame with the predicted SC of ORF1. TRXPT\_3,
similar to TRXPT\_1 but lacking the second exon, encodes a 13.1 kDa, 115
residue protein (ORF4), previously predicted (27) but excluded in later
annotations (1, 12). Our data suggest it is genuinely expressed. Lastly,
TRXPT\_4 encodes a distinct novel 15.9 kDa, 143 residue protein (ORF11).

The splice junctions of all transcripts in this region, except for
TRXPT\_4, were validated by cloning of viral cDNA and Sanger sequencing
(see \textbf{Supplementary PCR methods}). During TRXPT\_2 validation,
ORF1 was found on the agarose gel (an unspliced band size) and Sanger
sequencing results showed it to be a transcribed mRNA
(\textbf{Supplementary PCR methods}). This was corroborated by our 3'
RACE experiment, which showed a transcript (TRXPT\_2B) spanning the
entire ORF1 and Hyd ORFs without splicing, with a poly-A tail
immediately after the E1 TTS. The 5'-most coding sequence (CDS) of this
transcript (TRXPT\_2B) encodes ORF1. However, TRXPT\_2B has an upstream
and in-frame SC to the predicted SC of ORF1, suggesting that the
predicted ORF1 CDS is truncated -- the expressed ORF1 (eORF1) shares the
same SC as ORF10, but has a unique stop codon (STC). See
\textbf{Supplementary Table S3} for all transcripts and their encoded
proteins.

\textbf{Early Region 2 (E2) and Intermediate Region (IM) transcripts}\\
The E2 TU expressed on the anti-sense strand, is subdivided into E2A and
E2B and encodes three classical AdV proteins -- pTP and Ad-pol (E2B
proteins), and DBP (E2A protein) -- essential for genome replication
(17, 18). Unlike MAdV where two promoters are known (17), we discovered
only a single TSS (E2 TSS; 18,751 bp) for both E2A and E2B transcripts
in THEV. However, similar to MAdVs, E2A and E2B transcripts have
distinct TTSs, with E2B transcripts sharing the TTS of the IVa2
transcript of the IM region (17, 18) (\textbf{Figure 7}).

The E2A ORF, DBP, is one of three THEV ORFs predicted to be spliced from
two exons. The corresponding transcript (TRXPT\_21) in our data matches
the predicted splice junction but includes an additional non-coding exon
at the 5'-end (E2-5'UTR). Thus, TRXPT\_21 is a three-exon transcript
encoding DBP (380 residues, 43.3 kDa) precisely. TRXPT\_21 was also
corroborated in a 3' RACE experiment. Additionally, from the 3' RACE
data, we found a splice variant of TRXPT\_21 which retains the second
intron, leading to a 2-exon transcript (TRXPT\_21B). Although longer,
TRXPT\_21B encodes a truncated isoform of DBP (tDBP; a 346 residue, 39.3
kDa product) using a downstream in-frame SC but the same STC as DBP.
Both TRXPT\_21 and TRXPT\_21B share a common TTS, seen in our 3' RACE
data, located 39 bp downstream of the CDS in an adenine/thymine
(A/T)-rich sequence followed by the poly-A tail sequence, suggesting
this position (16,934 bp) as the true E2A TTS. There are two canonical
PASs (AAUAAA; 16,964 and 16950 bp) immediately after the CDS any of
which can serve as the PAS without affecting the encoded proteins
(\textbf{Supplementary Table S2}).

The E2B region transcripts also start with the E2-5'UTR but extend
downstream to reach the TTS at 2334 bp in the IM region, which is in the
immediate context of a canonical PAS (position 2333 bp) where
polyadenylation likely occurs. The E2B transcripts, TRXPT\_6 and
TRXPT\_7, are almost identical except for an extra splice junction at
the 3'-end of TRXPT\_6 (\textbf{Figure 7}). TRXPT\_7 has the CP for both
classical proteins (pTP and Ad-pol) encoded in this region, with the pTP
ORF predicted to be spliced from two exons. The predicted splice
junction of pTP is corroborated by our data but the full transcript is
markedly longer than the predicted ORF, although, the encoded product
(pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of
pTP with secondary SC (secSC) usage. The CP of TRXPT\_6 slightly differs
from TRXPT\_7 because a new STC resulting from the extra splice site
forms a minimal truncation of the Ad-pol encoded from its secSC.

While both TRXPT\_6 and TRXPT\_7 have the CP for Ad-pol with secSC
usage, in all AdVs studied, the two proteins (pTP and Ad-pol) are
encoded by separate mRNAs with identical first three 5' exons and TTS,
but different splice junctions to the terminal coding exons. Hence, we
checked for a longer splice junction between the third and fourth
(terminal) exons of TRXPT\_7 with our junction validation method
(targeted PCR, cloning, and Sanger sequencing). We discovered a unique
splice junction (10,981-7062 bp) not present in our RNA-seq data. If
initiated from the E2 TSS and terminated at the E2B TTS, this transcript
(TRXPT\_31) would encode Ad-pol in its 5'-most CDS (\textbf{Figure 7}).

@@

\textbf{Supplementary Table S2}

\begin{longtable}[c]{|p{0.80in}|p{0.80in}|p{2.00in}|p{0.80in}|p{1.50in}}
